---
title: "Modeling delay-discounting indifference points with bounded error distributions"
author: "Brent Kaplan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling delay-discounting indifference points with bounded error distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--
DRAFT (lives in dev/, not built). Becomes vignettes/sltb-discounting.Rmd once
fit_dd_tmb()/emmeans land and the de-duplicated Jarvis (2019) fixture is bundled
(data(jarvis2019)). Conceptual + boundary-demo chunks (1-3, 6) run today; the
API-usage chunks (4-5) are marked eval=FALSE until the functions exist. Apply
Brent's manuscript voice on the final pass.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
set.seed(1)
```

## Why indifference points need a bounded error distribution

A delay-discounting task asks a person to trade a smaller-sooner reward against a
larger-later one across a range of delays. The **indifference point** (IP) at each delay is
the smaller-sooner amount that feels equivalent to the later reward, expressed as a
*proportion of that later reward*. By construction an IP lives on the closed interval
$[0, 1]$:

- $y = 1$ — the person is indifferent only when the immediate amount equals the full later
  reward: **no discounting**.
- $y = 0$ — any immediate amount, however small, is preferred: **complete discounting**.

Real data pile up at both ends. In the Jarvis et al. (2019) data shipped with this package,
a meaningful fraction of observations sit *exactly* at 0 or 1. That single fact rules out
the two error distributions people reach for first:

- **Gaussian / nonlinear least squares** treats the residuals as unbounded and
  homoscedastic. It will happily predict an IP below 0 or above 1, and it assumes the noise
  is the same size near the bounds as in the middle — which is never true for a proportion.
- **Standard beta regression** is the natural model for a proportion, but its density is
  *undefined* at exactly 0 and 1 (the log-likelihood is $-\infty$). Every boundary
  observation has to be discarded or nudged, which biases the fit.

The **Scale-Location-Truncated beta (SLT-beta)** distribution (Kim, Koffarnus & Franck, 2024;
Kim, Kaplan, Koffarnus & Franck, 2025) keeps the beta's natural fit to proportion data but
assigns *finite probability* to the endpoints, so 0 and 1 are modeled rather than thrown away.

## What the SLT-beta density is

Write the mean of the IP at delay $D$ as a discounting function of a single rate $k$ — Mazur's
hyperbola $\mu = 1/(1 + kD)$ or the exponential $\mu = e^{-kD}$ — and let $\phi$ be a
precision (larger $\phi$ = tighter around the curve). With shape parameters
$a = \mu\phi$ and $b = (1-\mu)\phi$, and two *fixed* micro-constants $s = 1{+}10^{-7}$ and
$l = 10^{-8}$, the log-density of an observed IP $y \in [0,1]$ is

$$
\log f(y \mid \mu, \phi) =
\underbrace{\log\Gamma(a{+}b) - \log\Gamma(a) - \log\Gamma(b)}_{-\log B(a,b)}
+ (a{-}1)\log\!\big(\tfrac{y}{s}{+}l\big)
+ (b{-}1)\log\!\big(1 - \tfrac{y}{s}{+}l\big)
- \log s
- \log Z,
$$

where the truncation normalizer is $Z = \mathrm{pbeta}(\tfrac{1}{s}{+}l,\,a,\,b) -
\mathrm{pbeta}(l,\,a,\,b)$.

The intuition is a *microscopic stretch*: the data window $y\in[0,1]$ is mapped just inside
the open interval where the ordinary beta density is finite, so the kernel is never evaluated
at exactly 0 or 1. As $s\to 1$ and $l\to 0$ the SLT-beta collapses to the ordinary beta.

```{r density, fig.cap = "SLT-beta (solid) vs. ordinary beta (dashed). The SLT-beta carries finite density to y = 0 and y = 1; the ordinary beta diverges or vanishes there."}
slt_logpdf <- function(y, mu, phi, s = 1.0000001, l = 1e-8) {
  a <- mu * phi; b <- (1 - mu) * phi
  lgamma(a + b) - lgamma(a) - lgamma(b) +
    (a - 1) * log(y / s + l) + (b - 1) * log(1 - (y / s + l)) -
    log(s) - log(pbeta(1 / s + l, a, b) - pbeta(l, a, b))
}
yy <- seq(0, 1, length.out = 400)
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (p in list(c(mu = 0.5, phi = 4), c(mu = 0.15, phi = 6))) {
  plot(yy, exp(slt_logpdf(yy, p["mu"], p["phi"])), type = "l", lwd = 2,
       xlab = "indifference point", ylab = "density",
       main = sprintf("mu = %.2f, phi = %g", p["mu"], p["phi"]))
  lines(yy, dbeta(yy, p["mu"] * p["phi"], (1 - p["mu"]) * p["phi"]), lty = 2)
  points(c(0, 1), exp(slt_logpdf(c(0, 1), p["mu"], p["phi"])), pch = 16)
}
par(op)
```

Because the variance of a beta is $\mu(1-\mu)/(\phi+1)$, the SLT-beta automatically gives
**less** noise near the floor and ceiling and **more** in the middle — the heteroscedastic
pattern real IP data actually show, and exactly what a Gaussian model gets wrong.

## The boundary problem, on real data

```{r load-data}
## Final vignette: data(jarvis2019). Draft: read the CSV if present.
csv <- "~/Downloads/Data_brent.csv"
if (file.exists(path.expand(csv))) {
  d <- read.csv(path.expand(csv), check.names = FALSE)
  delays <- c(7, 30, 180, 365, 730, 1460, 2920)            # 1wk .. 8yr, in days
  long <- data.frame(id = rep(seq_len(nrow(d)), each = 7), # row = subject unit
                     delay = rep(delays, nrow(d)),
                     y = as.numeric(t(as.matrix(d[, -1]))))
  knitr::kable(as.data.frame(table(IP = long$y)), caption = "Observed indifference points")
}
```

The titration grid produces many observations at exactly 0 and 1. Counting subjects with at
least one boundary value, and asking which error models can even evaluate them:

```{r boundary-demo}
beta_logpdf <- function(y, mu, phi) {          # ordinary beta: -Inf at y in {0,1}
  a <- mu * phi; b <- (1 - mu) * phi
  lgamma(a + b) - lgamma(a) - lgamma(b) + (a - 1) * log(y) + (b - 1) * log(1 - y)
}
ids <- unique(long$id)
bnd <- sapply(ids, function(i) any(long$y[long$id == i] %in% c(0, 1)))
beta_ok <- sapply(ids, function(i) all(is.finite(beta_logpdf(long$y[long$id == i], 0.5, 8))))
slt_ok  <- sapply(ids, function(i) all(is.finite(slt_logpdf(long$y[long$id == i], 0.5, 8))))
data.frame(
  subjects                       = length(ids),
  with_boundary_IP               = sum(bnd),
  evaluable_under_ordinary_beta  = sum(beta_ok),
  evaluable_under_SLT_beta       = sum(slt_ok)
)
```

Every subject with a boundary value is lost to ordinary beta regression and retained by the
SLT-beta. Fitting each subject (estimating $\log k$ and $\log\phi$ so both stay positive) and
comparing the recovered rate to nonlinear least squares shows the SLT-beta agrees with NLS on
the bulk of subjects *and* succeeds on the boundary subjects NLS-plus-beta could not model:

```{r per-subject-fit}
fit_slt <- function(y, D) {                      # per-subject SLT MLE (log k, log phi)
  nll <- function(th) {
    mu <- pmin(pmax(1 / (1 + exp(th[1]) * D), 1e-6), 1 - 1e-6)
    v <- -sum(slt_logpdf(y, mu, exp(th[2]))); if (is.finite(v)) v else 1e10
  }
  o <- optim(c(log(0.01), log(8)), nll, control = list(reltol = 1e-11))
  exp(o$par[1])
}
nls_k <- function(y, D) {                        # multi-start NLS (single start is fragile)
  for (st in c(exp(-10), 1e-3, 1e-2, 1e-1)) {
    m <- tryCatch(nls(y ~ 1 / (1 + k * D), start = list(k = st)), error = function(e) NULL)
    if (!is.null(m)) return(coef(m)[["k"]])
  }
  NA_real_
}
k_slt <- sapply(ids, function(i) fit_slt(long$y[long$id == i], long$delay[long$id == i]))
k_nls <- sapply(ids, function(i) nls_k(long$y[long$id == i], long$delay[long$id == i]))
sprintf("cor(log k): SLT vs NLS = %.3f   |   median k: SLT = %.4f, NLS = %.4f",
        cor(log(k_slt), log(k_nls), use = "complete.obs"),
        median(k_slt), median(k_nls, na.rm = TRUE))
```

A caveat the next section resolves: fit *one subject at a time* and the SLT-beta likelihood
can occasionally run off to a degenerate solution — for a person whose indifference points are
almost all 0s and 1s, a fit with near-zero precision that places all its mass on the endpoints
can score a higher likelihood than any sensible discounting curve. The fix is not to fit
people in isolation.

## Fitting the mixed-effects model

<!-- eval=FALSE until fit_dd_tmb() lands; this is the intended API. -->

A mixed-effects model shares information across people. The subject random intercept on
$\log k$ acts as a prior that pulls each person's estimate toward the population, so the
degenerate per-subject solutions above are regularized away — one of the main reasons to
prefer `fit_dd_tmb()` over fitting each subject separately.

In practice you rarely want a separate $k$ per person in isolation. `fit_dd_tmb()` fits a
**mixed-effects** discounting model: a population rate with a subject random intercept on
$\log k$, plus any between-subject factors, all under the SLT-beta likelihood.

```{r fit-tmb, eval = FALSE}
fit <- fit_dd_tmb(
  data      = jarvis2019,            # columns: id, x (delay), y (indifference point)
  equation  = "mazur",               # or "exponential"
  family    = "sltb",                # default; "gaussian" reproduces the NLS-style fit
  random_effects = k ~ 1             # random intercept on log k (fit internally on log scale)
)
summary(fit)
tidy(fit)                            # population log k, precision; back-transformed k
ranef(fit)                           # per-subject discount rates
```

Group contrasts use the same estimated-marginal-means surface as `beezdemand`, computed on
the `log k` scale and back-transformed, so a between-group comparison reads as a **ratio of
discount rates**:

```{r emmeans, eval = FALSE}
fit_grp <- fit_dd_tmb(jarvis2019, equation = "mazur", factors = "group")
get_dd_param_emms(fit_grp)                       # EMM of k per group
get_dd_comparisons(fit_grp, ~ group, adjust = "holm")  # ratio-of-k contrasts
```

## Choosing a family

| | `family = "sltb"` (default) | `family = "gaussian"` |
|---|---|---|
| Respects $[0,1]$ | yes (incl. boundary mass at 0 and 1) | no (can predict outside) |
| Variance | shrinks near bounds, grows mid-range | constant |
| Boundary data | modeled | only via the bounded mean |
| Matches legacy `fit_dd()` | — | yes (NLS-style) |

Use `"sltb"` for inference on real IP data; use `"gaussian"` when you want continuity with a
classic least-squares discounting analysis or a quick baseline.

## How much to trust it

The SLT-beta density was checked before release: it integrates to one, has finite density at
both boundaries, reproduces the published moments, and recovers known discount rates from
simulated data (rank correlation 0.99 for Mazur, 0.95 for the exponential). On the Jarvis
(2019) data every boundary subject is fit successfully where ordinary beta regression cannot
be evaluated at all. The full Monte-Carlo study lives in the package's `dev/` directory.

## References

- Kim, M., Koffarnus, M. N., & Franck, C. T. (2024). *Thinking Inside the Bounds: Improved
  Error Distributions for Indifference Point Data Analysis and Simulation via Beta Regression
  Using Common Discounting Functions.*
- Kim, M., Kaplan, B. A., Koffarnus, M. N., & Franck, C. T. (2025). *Scale-Location-Truncated
  Beta Regression: Expanding Beta Regression to Accommodate 0 and 1.* arXiv:2509.13167.
- Jarvis, B. P., et al. (2019). Delay-discounting dataset (N = 126; 1 week – 8 years).
