---
title: "Bayesian discounting models with brms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian discounting models with brms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      eval = FALSE, purl = FALSE)
```

The Bayesian tier fits the same discounting models as `fit_dd_tmb()` and
`fit_dd_choice(mode = "structural")` using brms and Stan, with the discount
rate estimated on the same natural-log scale. Nothing in this vignette runs
at build time (Stan sampling takes minutes); copy the chunks into your
session. brms, posterior, and loo are suggested packages.

## Indifference points

```{r fit, eval = FALSE}
library(beezdiscounting)
data(dd_ip)

fit <- fit_dd_brms(
  dd_ip,
  equation = "mazur",
  family = "beta",
  chains = 4, cores = 4, seed = 1
)

summary(fit)
tidy(fit)        # k on the natural scale (posterior medians of exp draws)
confint(fit)     # equal-tailed credible intervals
ranef(fit)       # per-subject k posterior summaries
```

### Families and boundaries

The TMB tier's SLT-beta family has no brms equivalent. Two options cover
the same ground:

- `family = "beta"` (default): brms's beta likelihood with an identity
  link; the mean function is squished into (1e-6, 1 - 1e-6), the
  differentiable analog of the TMB clamp. Boundary responses (`y` exactly
  0 or 1) are handled by `boundary`: the default `"squeeze"` applies the
  Smithson-Verkuilen transform `(y(N-1) + 0.5)/N` to all responses and
  reports the boundary count; `"zoib"` switches to a zero-one-inflated
  beta -- statistically more explicit about boundary responding, but it
  **changes the estimand**: k then describes the interior responses only;
  `"error"` refuses to fit.
- `family = "gaussian"`: matches `fit_dd_tmb(family = "gaussian")`
  wherever the TMB mean clamp (into `[1e-6, 1 - 1e-6]`) does not bind --
  everywhere except extreme decay underflow. In our validation harness the
  posterior median of log k lands within hundredths of the TMB MLE.

### Priors

k is delay-unit dependent (per-day vs per-week k differ by a factor of 7),
so the default `logk` prior anchors to the data: it centers
`k * median(delay) = 1`, the delay at which the Mazur curve crosses one
half. The anchor is printed at fit time, stored in `fit$autoscale_info`,
and the whole table is inspectable and overridable:

```{r priors, eval = FALSE}
default_dd_priors("mazur", family = "beta", data = dd_ip)

fit2 <- fit_dd_brms(
  dd_ip,
  equation = "mazur",
  prior = brms::set_prior("normal(-3, 1)", class = "b",
                          coef = "Intercept", nlpar = "logk")
)
```

The two-parameter equations (`"green-myerson"`, `"rachlin"`) add a
population-level shape exponent with prior `normal(0, 0.5)` on log s
(s near 1 a priori); zero delays are valid for Rachlin (the implementation
guards the `0^s` case exactly as the TMB template does).

## Trial-level choice

```{r choice, eval = FALSE}
fitc <- fit_dd_choice_brms(
  choice_data,
  equation = "mazur",
  chains = 4, cores = 4, seed = 1
)

tidy(fitc)              # k and gamma (choice sensitivity), natural scale
predict(fitc)           # P(choosing the larger-later), with intervals
```

The structural likelihood is identical to
`fit_dd_choice(mode = "structural")`:
`logit P(LL) = [b0] + gamma ((ll/ss) D(k, delay) - 1)`. The descriptive
(Young 2018) mode is not wrapped -- it is a plain logistic GLMM you can
fit directly with `brms::brm()` if you want it Bayesian.

## Reading Bayesian output

- `tidy()` keeps the package's exact eight-column contract, with
  `statistic`/`p.value` set to `NA`: estimates are posterior medians of the
  report-space-transformed draws (no delta method), and intervals live in
  `confint()`.
- `glance()` reports `elpd_loo`/`looic` plus R-hat/ESS/divergence
  diagnostics; `logLik`/`AIC`/`BIC` are `NA` by design. Compare Bayesian
  fits with `brms::loo_compare(fit$brmsfit, fit2$brmsfit)`.
- `fit$converged` requires R-hat < 1.01, zero divergences, and bulk
  ESS >= 400; components are in `fit$mcmc_info`, and the full brms
  toolbox is available on `fit$brmsfit`.

## v1 scope

Random effects are intercept-only on k (`k ~ 1`); the TMB tier's
`k + phi ~ 1` is planned via a distributional formula. Bayesian
EMMs/comparisons for factor designs are deferred to a follow-up release.
