---
title: "Comparing discounting rates between groups"
author: "Brent Kaplan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing discounting rates between groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
has_tmb <- requireNamespace("TMB", quietly = TRUE)
```

A common discounting question is not "what is k?" but "does k differ
between groups?" -- treatment vs. control, clinical vs. community,
commodity A vs. commodity B. Both modeling tiers in beezdiscounting answer
it the same way: the discount rate is estimated on the natural-log scale,
so a between-subject design enters as a linear model on log k
(`log k = X beta`), and every marginal mean or contrast is a linear
combination of coefficients. `get_dd_param_emms()` and
`get_dd_comparisons()` work on top of that shared backbone for:

- `fit_dd_tmb()` -- mixed-effects indifference-point models (Wald
  inference),
- `fit_dd_choice()` -- the structural trial-level choice model (Wald
  inference),
- `fit_dd_brms()` and `fit_dd_choice_brms()` -- their Bayesian
  counterparts (posterior draws; see the last section).

The TMB examples below run at build time; the Bayesian section shows code
without running it (Stan sampling takes minutes -- copy the chunks into
your session).

## Simulate a two-group study

`simulate_dd_ip()` generates indifference points with a known
between-condition shift in log k. Here condition `C2`'s k is three times
`C1`'s (`delta_k = log(3)`). To have a second factor for the grouped
contrasts shown later, we also assign each subject a (null) `site`.

```{r simulate, eval = has_tmb}
library(beezdiscounting)

sim <- simulate_dd_ip(
  n_subjects = 24, n_conditions = 2, delta_k = c(0, log(3)), seed = 42
)
# conditions alternate across subjects, so assign site in blocks of two to
# get the full condition x site crossing
subjects <- unique(sim$id)
sim$site <- factor(ifelse(
  ((match(sim$id, subjects) - 1) %/% 2) %% 2 == 0, "siteA", "siteB"
))
with(unique(sim[, c("id", "condition", "site")]), table(condition, site))
head(sim)
```

## Fit with a factor design on log k

`factors` puts the design on log k; every other parameter stays
population-level. (`multi_start = FALSE` keeps the vignette quick; leave
the default on for real analyses.)

```{r fit-tmb, eval = has_tmb}
fit <- fit_dd_tmb(
  sim,
  factors = c("condition", "site"),
  multi_start = FALSE, verbose = 0
)
tidy(fit)
```

## Estimated marginal means of k

`get_dd_param_emms()` averages the fitted design over the factor levels
you are not looking at (equal weights, the emmeans convention) and
back-transforms: `k = exp(k_log)`, with the standard error and the Wald
interval formed on the log scale.

```{r emms, eval = has_tmb}
get_dd_param_emms(fit)
```

Marginalize over `site` to get per-condition means, or condition on a
level with `at`:

```{r emms-marginal, eval = has_tmb}
get_dd_param_emms(fit, factors_in_emm = "condition")

get_dd_param_emms(fit, at = list(site = "siteA"))
```

## Contrasts

`get_dd_comparisons()` turns the same reference grid into factor-level
contrasts. Estimates are reported on the log10 scale (the customary
discounting metric) and, because differences in log k are log ratios,
also as multiplicative ratios: a ratio of 3 means one group discounts
three times as steeply. P-values are adjusted with `stats::p.adjust()`
(`"holm"` by default).

```{r contrasts, eval = has_tmb}
cmp <- get_dd_comparisons(fit, compare_specs = ~condition)
cmp$k$contrasts_log10
cmp$k$contrasts_ratio
```

The simulated truth was `k(C2) = 3 k(C1)`, so the displayed `C1 - C2`
ratio should land near 1/3 -- and it does. With more than one factor,
`contrast_by` conditions the contrasts on another factor's levels
(adjustment is applied per by-cell; with no interaction in the model the
per-site contrasts coincide, as here):

```{r contrast-by, eval = has_tmb}
cmp_by <- get_dd_comparisons(
  fit,
  compare_specs = ~ condition + site, contrast_by = "site"
)
cmp_by$k$contrasts_log10
```

`tidy()` flattens any comparison object into one cross-backend frame:

```{r tidy-cmp, eval = has_tmb}
tidy(cmp)
```

## Trial-level choice models

The structural choice model estimates the same log-k design from binary
smaller-sooner vs. larger-later choices, so the comparison surface is
identical. (The simulator has no built-in group effect; stack two
simulated groups.)

```{r choice, eval = has_tmb}
ctrl <- simulate_dd_choice(n_subjects = 12, log_k_pop = log(0.01), seed = 7)
trt <- simulate_dd_choice(n_subjects = 12, log_k_pop = log(0.04), seed = 8)
trt$id <- paste0("t", trt$id)
ctrl$group <- "ctrl"
trt$group <- "treat"
choice_data <- rbind(ctrl, trt)

fit_choice <- fit_dd_choice(
  choice_data,
  mode = "structural", factors = "group", verbose = 0
)
get_dd_param_emms(fit_choice)
get_dd_comparisons(fit_choice)$k$contrasts_ratio
```

## The Bayesian analogs

`fit_dd_brms()` and `fit_dd_choice_brms()` take the same
`factors`/`factor_interaction`/`continuous_covariates` arguments, and
`get_dd_param_emms()`/`get_dd_comparisons()` accept the fits directly:

```{r brms, eval = FALSE}
fit_b <- fit_dd_brms(
  sim,
  equation = "mazur", factors = c("condition", "site"),
  chains = 4, cores = 4, seed = 1
)
get_dd_param_emms(fit_b, factors_in_emm = "condition")
get_dd_comparisons(fit_b, compare_specs = ~condition)

fit_cb <- fit_dd_choice_brms(
  choice_data,
  equation = "mazur", factors = "group",
  chains = 4, cores = 4, seed = 1
)
get_dd_comparisons(fit_cb)
```

The tables have the same shape, with draws-based summaries in place of
Wald machinery:

- Estimates are posterior medians; intervals are equal-tailed quantiles
  of the (transformed) draws -- the marginal means and contrasts are
  computed per posterior draw over the same reference grid, then
  summarized, so back-transformation is exact (no delta method).
- `statistic`, `df`, and `p.value` are `NA`. In their place `post.prob`
  reports the posterior probability of direction: the probability that
  the contrast is positive (or negative, whichever is larger). A
  `post.prob` near 1 plays the practical role of a small p-value.
- There is no multiplicity adjustment, by design: the contrasts are
  deterministic functions of one joint posterior, which already encodes
  their dependence. Asking for `adjust` produces a warning rather than an
  adjusted column.

## Which backend when

Use the TMB fitters for fast iteration and conventional Wald tests with
familiar adjusted p-values. Reach for the brms fitters when you want
full posterior uncertainty for derived quantities (the EMM and ratio
intervals above need no asymptotics), prior information about plausible
discounting rates, or small-sample inference you can defend without
appealing to asymptotic normality. The vignette
`vignette("bayesian-discounting")` covers the Bayesian tier itself --
priors, families, convergence -- in more depth.
