---
title: "The linearized Mazur hyperbola: closed-form k and an F-test for conditions"
author: "Brent Kaplan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The linearized Mazur hyperbola: closed-form k and an F-test for conditions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
set.seed(1)
library(beezdiscounting)
```

The other modeling tiers in `beezdiscounting` estimate the discount rate by
optimization. `fit_dd()` fits Mazur's hyperbola to each subject by nonlinear least
squares, which may fail to converge for flat or erratic subjects, and `fit_dd_tmb()`
fits a mixed model whose marginal likelihood has to be maximized numerically. Hinds
et al. (2026) take a different route. They rearrange the hyperbola so that ln k
appears as an intercept, at which point the per-subject estimate is a mean, the
population model is a one-way random-effects ANOVA with closed-form maximum
likelihood estimates, and condition means can be compared with an F-test whose
null distribution holds at any sample size under the model's assumptions. No optimizer is
involved anywhere, so there is nothing to fail to converge. `fit_dd_linear()`
implements that estimator.

This vignette covers (i) the transform and the per-subject estimator, (ii) the
random-effects model and the methods that act on a fit, (iii) comparing conditions
with the F-test, (iv) what to do with indifference points at exactly 0 or 1, and
(v) how the linearized estimates relate to the nonlinear ones.

## The transform

Mazur's (1987) hyperbola gives the discounted value of a delayed reward as a
proportion of its undelayed value, `D = 1 / (1 + k t)`, where `t` is the delay and
`k` is the discount rate. Solving for `k` and taking logs,

```
ln(1 / D - 1) - ln(t) = ln k .
```

Every indifference point, once transformed this way, is a direct read of ln k. Hinds
et al. (2026) treat the transformed points as ln k plus additive Gaussian error, so the
per-subject estimate of ln k is the mean of the transformed points (equivalently, the
geometric mean of `(1 / D - 1) / t`). Note that the Gaussian error on the transformed
scale is a modeling assumption, as it is for any other error model; Section 2 of the
paper discusses it, and the last section below shows what it implies on the raw scale.

To make that concrete, here is one subject drawn from the simulator introduced in
the next section, with the transform done by hand:

```{r by-hand}
one <- simulate_dd_linear(1, delays = c(7, 30, 90, 180, 365, 730),
                          mu = c(demo = -6), sigma2 = 2, g = 10, seed = 3)
one$y_lin <- log(1 / one$y - 1) - log(one$x)
one
mean(one$y_lin)
```

`fit_dd_linear()` returns the same number:

```{r by-hand-fit}
fit_dd_linear(one)$subjects[, c("id", "logk", "se", "ci_lo", "ci_hi")]
```

The standard error and interval are the usual t-based quantities for a mean of
`T` observations (here `T = 6` delays), so for a given spread of transformed points
a subject measured at more delays gets a narrower interval.

## Simulate a three-condition study

`simulate_dd_linear()` draws data from the random-effects model that
`fit_dd_linear()` fits: each subject's ln k is drawn around their condition mean
`mu`, and each transformed point is that subject's ln k plus Gaussian noise. The
model has two variance parameters. `sigma2` is the noise variance on the transformed
scale, and `g` scales the between-subject variance, `Var(ln k_i) = g * sigma2 / T`.
The values below are of the order the paper reports for real data
(`sigma2` near 2, `g` near 10). A scalar `n_subjects` is recycled across
conditions, so this call yields 30 subjects per condition and 90 in total.
Conditions `A` and `C` share a mean; `B` discounts more steeply by 1.5 on the ln k
scale (a factor of about 4.5 in `k`).

```{r sim}
sim <- simulate_dd_linear(
  n_subjects = 30,
  delays     = c(7, 30, 90, 180, 365, 730),
  mu         = c(A = -6, B = -4.5, C = -6),
  sigma2     = 2,
  g          = 10,
  seed       = 1
)
head(sim)
```

The layout is the package's usual long format (`id`, `x` for delay, `y` for the
indifference point on `[0, 1]`) plus the `condition` column.

## Fit and inspect

`factors` names the one between-subject column that defines the conditions. Two
design requirements apply to the random-effects part of the fit: every subject must
appear under exactly one level (the F-test assumes independent units per condition),
and every subject must contribute the same number of usable delays. A within-subject
factor is rejected with an error; an unbalanced design (see the boundary section
below) returns the per-subject estimates but leaves the population component
unfitted.

```{r fit}
fit <- fit_dd_linear(sim, factors = "condition")
fit
```

The print method reports the design, the boundary bookkeeping (nothing to report
here, as the simulator keeps every point strictly inside (0, 1)), the condition means
`mu` on the ln k scale, and the two variance estimates. `summary()` adds the first
rows of the per-subject table and the population parameters as a tibble:

```{r summary}
summary(fit)
```

`coef()` returns the condition means and `confint()` an interval around each.
The paper gives no population interval, so the package uses the ANOVA-style
t-interval on the per-subject ln k values (between-subject mean square,
`N - C` degrees of freedom); with `parm = "subject"` you get the per-subject
intervals instead.

```{r coef}
coef(fit)
confint(fit)
head(confint(fit, parm = "subject"), 3)
```

`tidy()` returns the per-subject table by default (`logk`, its `se` and interval,
`k` back-transformed with its interval, and `n_boundary`, the count of points that
sat at 0 or 1), or the population parameters with `effects = "population"`:

```{r tidy}
tidy(fit)
tidy(fit, effects = "population")
```

Because the data were simulated, the estimates can be set beside the values that
generated them:

```{r recovery}
tibble::tibble(
  term     = c("mu_A", "mu_B", "mu_C", "sigma2", "g"),
  truth    = c(-6, -4.5, -6, 2, 10),
  estimate = tidy(fit, effects = "population")$estimate
)
```

The means and `sigma2` are recovered closely; `g` is estimated with more
uncertainty, as a between-subject variance from 30 subjects per condition would be
under any estimator.

`glance()` gives the one-row model summary used across the package for stacking
fits (`nobs()` returns the count of usable transformed points, 540 here), and
`logLik()` the log-likelihood. By default `logLik()` returns the
population (random-effects) likelihood on the raw indifference-point scale, that
is, corrected by the Jacobian of the transform (paper Section 2.3), which is what
makes it comparable with the Gaussian likelihoods of `fit_dd()` and
`fit_dd_tmb()`. `scale = "transformed"` returns the likelihood of the transformed
points, and `level = "subject"` the sum of the per-subject likelihoods (the
quantity to set against a per-subject fit rather than a mixed model).

```{r glance}
glance(fit)
logLik(fit)
```

`augment()` returns the long data with the transformed value `y_lin`, the value
actually used after boundary handling `d_used`, the fitted indifference point
and residual on the raw scale, `.fitted` and `.resid`, and the standardized
residual on the transformed scale, `.std_resid` (the residual divided by the
model's error standard deviation, `sqrt(sigma2)`). Together with `tidy()`,
`coef()` and `confint()`, this table is what you need to reproduce or customize
the figures that `plot()` draws in the Figures section.

```{r augment}
head(augment(fit), 3)
```

## Comparing conditions

`anova()` tests whether the condition means are equal. The statistic is the
paper's F (Proposition 3.5) with `C - 1` and `N - C` degrees of freedom (here 2
and 87). Under the model's assumptions (a balanced design, independent subjects,
and Gaussian error on the transformed scale) its null distribution is F at any
`N`, with no large-sample approximation involved. The proposition derives this
for a positive between-subject variance estimate `g`, which is the usual case;
when `g` is estimated at 0 the package still reports the statistic but warns.

```{r anova}
anova(fit)
```

A `hypothesis` argument restricts the null to a subset of levels. To test only
whether `A` and `C` share a mean (the two conditions the simulation set equal):

```{r anova-hyp}
anova(fit, hypothesis = list(c("A", "C")))
```

`pairwise = TRUE` runs every two-level comparison. The p-values are uncorrected,
and each row also reports Cohen's d on the ln k scale (the mean difference divided
by the between-subject standard deviation); `cohens_d` is only defined for
two-level hypotheses, which is why the overall test above shows `NA`.

```{r anova-pairwise}
anova(fit, pairwise = TRUE)
```

`B` differs from both `A` and `C` at about one between-subject standard deviation,
and `A` and `C` do not differ, which is the structure the simulation built in.
By contrast, the Wald tests in `vignette("dd-group-comparisons")` rely on a
large-sample approximation to the sampling distribution. The F-test here needs
no such approximation, and it costs nothing beyond the closed-form fit.

## Indifference points at 0 or 1

The transform is undefined at `D = 0` and `D = 1` (`ln(1 / D - 1)` diverges), and
the paper does not address such values. Real titration data contain them. The
package's `dd_ip` dataset has 600 points from 100 subjects, of which 125 are exactly
0 and 4 lie above 1 (the input check clamps those to 1, with a warning), so 129
points sit on the boundary. `boundary` selects one of three policies.

**`"clamp"`** (the default) moves an exact 0 to `eps` and an exact 1 to `1 - eps`.
Only exact 0 and 1 are touched; an interior point, however close to a bound, is
used as observed. The print method reports how many points were clamped.

```{r dd-ip-clamp}
data(dd_ip)
fit_clamp <- fit_dd_linear(dd_ip)
fit_clamp
```

**`"drop"`** removes boundary points. The per-subject estimates are still returned,
but subjects then differ in their number of usable delays, the design is no longer
balanced, and the random-effects component is unavailable. The population methods
(`coef()`, `anova()`, and the default `confint()`, `tidy(effects = "population")`
and `logLik()`) then error; the per-subject versions still work.

```{r dd-ip-drop}
fit_drop <- fit_dd_linear(dd_ip, boundary = "drop")
fit_drop
```

**`"error"`** refuses the data, which is the right choice in a pipeline that should
never silently impute:

```{r dd-ip-error, error = TRUE}
fit_dd_linear(dd_ip, boundary = "error")
```

With a fifth of the points clamped, the population estimate depends on where the
clamped points are placed. The default `eps` is 0.005; when `ll` (the larger-later
amount) is supplied it becomes `1 / (2 * ll)`, half the smallest step the task
could resolve. The table below refits `dd_ip` across a range of `eps`:

```{r eps-sensitivity, warning = FALSE}
eps_grid <- c(0.001, 0.005, 0.01, 0.05)
tibble::tibble(
  eps = eps_grid,
  mu  = sapply(eps_grid, function(e) coef(fit_dd_linear(dd_ip, eps = e))),
  k   = exp(mu)
)
```

The estimate of ln k moves by 0.8 across this range, which is a factor of about 2.2
in `k`. The sensitivity arises when data with many points on the boundary meet a
transform that diverges there and a clamp policy that has to pick a finite
stand-in. A bounded likelihood needs no `eps` at all. We recommend setting
`eps` from the task's resolution via `ll`, reporting the clamped count, and, when
boundary points are this common, preferring a likelihood that treats them as data;
see `vignette("sltb-discounting")`.

## Figures

`plot()` draws the pictures that go with the fit, with the same arguments as the
other indifference-point tiers (`type`, `ids`, `n_points`, `x_trans`,
`show_observed`). The linearization is easiest to see on the transformed scale.
For each subject, `ln(1 / D - 1)` plotted against `ln(t)` should scatter around a
line of slope 1 whose intercept is that subject's ln k. `type = "transformed"`
draws that picture, one panel per subject; `ids` picks the subjects (without it,
the first twelve are shown):

```{r fig-transformed, fig.alt = "Six panels, one per simulated subject, plotting ln(1/D - 1) against ln(delay); the points in each panel scatter around a line of slope 1 whose intercept is the subject's estimated ln k."}
show <- c("A_1", "A_2", "B_1", "B_2", "C_1", "C_2")
plot(fit, type = "transformed", ids = show)
```

Back on the raw scale, `plot(fit)` draws the hyperbola implied by each
condition's geometric-mean k, `exp(mu)`, over the observed indifference points.
`type = "individual"` adds each subject's own hyperbola, `1 / (1 + k x)`,
underneath (here for the same six subjects):

```{r fig-raw, fig.alt = "Observed indifference points against delay on a log axis, with three coloured condition curves and six thin per-subject hyperbolae drawn underneath them."}
plot(fit, type = "individual", ids = show)
```

The third figure answers the study question. `type = "parameters"` shows every
subject's k with its t-interval, by condition, with the condition's
geometric-mean k and its interval (the exponentiated values from `coef()` and
`confint()`) overlaid. The axis is k on a log scale; `coef()` and `confint()` report the same
quantities as ln k.

```{r fig-conditions, fig.alt = "Per-subject discount rates k with confidence intervals on a log axis, jittered within three conditions A, B, and C; condition B sits higher than A and C, and the condition geometric means with their intervals are overlaid in a contrasting colour."}
plot(fit, type = "parameters")
```

The frames behind these figures come from `augment()`, `tidy()`, `coef()` and
`confint()`, for anyone who wants a different picture.

## How this relates to the nonlinear fit

The same data fitted subject by subject with `fit_dd()` give a second set of
per-subject rates. The two estimators target the same `k` under different error
models: `fit_dd()` assumes Gaussian error on the indifference-point scale, whereas
`fit_dd_linear()` assumes it on the transformed scale, where a fixed error in
`ln(1 / D - 1)` corresponds to a small raw-scale error near `D = 0.5` and a much
smaller one near the bounds.

```{r nls-compare}
nls_fit <- fit_dd(sim, equation = "mazur", method = "two stage")
nls_res <- results_dd(nls_fit)
both <- merge(
  tidy(fit)[, c("id", "condition", "logk")],
  data.frame(id = nls_res$id, logk_nls = log(nls_res$estimate)),
  by = "id"
)
cor(both$logk, both$logk_nls)
head(both)
```

The two sets of ln k agree closely on these data (`r sprintf("r = %.2f", cor(both$logk, both$logk_nls))`),
and the largest disagreements are for subjects with a point very close to 0 or 1,
where the two error models weight the observations most differently. Which error
model is closer to the truth is an empirical question. Hinds et al. (2026) compare
the estimators by simulation and on real data, and we refer the reader there
rather than repeating the comparison. A likelihood comparison between the two is
possible because `logLik(fit)` is Jacobian-corrected to the indifference-point
scale, which puts it on the same scale as a Gaussian log-likelihood on the raw
points. Comparability also requires that (i) the same observations enter both
fits, (ii) boundary handling is identical, and (iii) the likelihood levels match
(`level = "subject"` against a per-subject fit, the default population likelihood
against a mixed model such as `fit_dd_tmb()`).

## Where to go next

- `vignette("dd-group-comparisons")` for factor designs, marginal means, and
  contrasts under the mixed-effects and Bayesian tiers.
- `vignette("sltb-discounting")` for a bounded likelihood that treats indifference
  points at 0 and 1 as data rather than as a nuisance.
- `vignette("tmb-mixed-effects")` for the `fit_dd_tmb()` workflow, whose
  log-likelihood is on the same scale as `logLik(fit)` here.

## References

- Hinds, D., Tegge, A. N., Stein, J. S., LaConte, S. M., McClure, S. M., & Ferreira,
  M. A. R. (2026). To linearize or not to linearize: That is the Mazur delay
  discounting question. *Journal of Mathematical Psychology, 130*, 103006.
  https://doi.org/10.1016/j.jmp.2026.103006
- Mazur, J. E. (1987). An adjusting procedure for studying delayed reinforcement.
  In *The effect of delay and of intervening events on reinforcement value*
  (pp. 55--73). Lawrence Erlbaum Associates.
