---
title: "Mixed-effects discounting with TMB"
author: "Brent Kaplan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed-effects discounting with TMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_tmb <- requireNamespace("TMB", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2,
                      eval = has_tmb)
set.seed(1)
library(beezdiscounting)
```

Fitting each subject's discounting curve in isolation ignores that subjects are
samples from a population. The mixed-effects tier in `beezdiscounting` estimates a
population discount rate and each subject's rate at the same time, borrowing strength
across subjects so that noisy individuals are pulled toward the group. `fit_dd_tmb()`
is the engine for that tier.

This vignette is a tour of the `fit_dd_tmb()` workflow and the methods that act on a
fitted model. It pairs with two other vignettes: `vignette("sltb-discounting")`
explains *why* indifference points need a bounded error distribution and develops the
scale-location-truncated beta (SLT-beta) likelihood used below, and
`vignette("dd-group-comparisons")` goes deeper on contrasting groups. Here the focus
is the model object and what you can do with it.

## Why TMB

The discount rate is estimated on the natural-log scale, where the log-k predictor
is linear in the fixed effects, which makes each subject's rate
`k_i = exp(population log-k + random deviation)`. Integrating the random deviations
out of the likelihood has no closed form. Template Model Builder (TMB) handles it
with a Laplace approximation and automatic differentiation: exact gradients, a fast
compiled objective, and maximum-likelihood estimates in seconds rather than minutes.
TMB must be installed for any of this to run.

## Some data

`simulate_dd_ip()` generates indifference-point data from the same model
`fit_dd_tmb()` fits, which makes it convenient for a reproducible walk-through. The
built-in `dd_ip` dataset works exactly the same way.

```{r sim}
sim <- simulate_dd_ip(
  n_subjects = 15,
  delays     = c(7, 30, 90, 180, 365, 730),
  log_k_pop  = log(0.01),
  sigma_u    = 0.5,
  phi        = 10,
  family     = "sltb",
  equation   = "mazur",
  seed       = 1
)
head(sim)
```

The data are long format with columns `id`, `x` (delay), and `y` (indifference
point on `[0, 1]`), the same layout used throughout the package.

## Fitting a model

A one-group hyperbolic model with a random intercept on log-k and the SLT-beta
likelihood:

```{r fit}
fit <- fit_dd_tmb(
  sim,
  equation       = "mazur",
  family         = "sltb",
  random_effects = k ~ 1
)
fit
```

The `random_effects = k ~ 1` formula puts a subject random intercept on the discount
rate. `family = "sltb"` selects the bounded error distribution; use
`family = "gaussian"` for unbounded residuals.

## Inspecting the fit

`summary()` reports the fixed effect (the population `k` on the natural scale), the
variance components, and fit statistics:

```{r summary}
summary(fit)
```

`tidy()` returns the same information as a data frame (one row per term, with the
scale each estimate is reported on), which is handy for tables and for stacking
models:

```{r tidy}
tidy(fit)
```

`glance()` gives a one-row model summary for comparisons:

```{r glance}
glance(fit)
```

## Subject-level rates

`ranef()` returns each subject's random deviation (`u_i`) and their fitted discount
rate (`k`):

```{r ranef}
head(ranef(fit))
```

These are the shrinkage estimates: subjects with noisy or extreme data are pulled
toward the population rate, more so when they have fewer or messier observations.

## Predictions

`predict()` returns fitted indifference points. The `level` argument chooses between
the population-average curve (random effects set to their mean) and subject-specific
curves. Predict the population curve on a fine grid of delays and plot it over the
raw points:

```{r predict-plot}
grid <- data.frame(x = seq(1, 730, length.out = 100))
head(predict(fit, newdata = grid, level = "population"))

# plot() draws the population curve over the observed points directly; type =
# "individual" adds the per-subject (shrinkage) curves.
plot(fit, type = "population")
```

For data that includes an `id` column, request both levels at once with
`level = c("population", "subject")` to get the population prediction and each
subject's prediction side by side.

## Diagnostics

`augment()` attaches fitted values and residuals to the model frame, including
standardized residuals that account for the SLT-beta's non-constant variance:

```{r augment}
aug <- augment(fit)
head(aug)

plot(fit, type = "resid")
```

## Choosing an equation

`fit_dd_tmb()` fits four discounting functions: `"mazur"` and `"exponential"`
(one-parameter), and `"green-myerson"` and `"rachlin"` (two-parameter hyperboloids
that add a curvature exponent). Fit several and compare them with `glance()`:

```{r compare-eq}
equations <- c("mazur", "exponential", "green-myerson")
fits <- lapply(equations, function(eq)
  fit_dd_tmb(sim, equation = eq, family = "sltb", random_effects = k ~ 1))

do.call(rbind, lapply(fits, function(f) glance(f)[c("equation", "logLik", "AIC", "BIC")]))
```

Because these data were generated from a hyperbola, Mazur wins on AIC; the
exponential fits worse, and the extra parameter in Green-Myerson does not earn its
keep. With real data the ranking is an empirical question worth checking.

## Beyond one group and one random effect

Add between-subject factors with the `factors` argument, then get per-group rates and
contrasts. Simulate two groups and fit a model with a condition effect on log-k:

```{r groups}
sim2 <- simulate_dd_ip(
  n_subjects = 30, delays = c(7, 30, 90, 180, 365, 730),
  log_k_pop = log(0.01), sigma_u = 0.6, phi = 12,
  family = "sltb", equation = "mazur",
  n_conditions = 2, delta_k = c(0, log(2)), seed = 42
)

fit2 <- fit_dd_tmb(sim2, equation = "mazur", family = "sltb",
                   random_effects = k ~ 1, factors = "condition")

get_dd_param_emms(fit2)
```

`get_dd_comparisons()` turns those marginal means into contrasts, reported both on
the log10 scale and as multiplicative ratios with Holm-adjusted p-values:

```{r comparisons}
cmp <- get_dd_comparisons(fit2, compare_specs = ~ condition,
                          contrast_type = "pairwise", adjust = "holm")
cmp$k$contrasts_ratio
```

See `vignette("dd-group-comparisons")` for the full treatment of group inference.

The random-effects structure can grow too. With the SLT-beta family,
`random_effects = k + phi ~ 1` adds a subject random effect on the precision, letting
each subject have their own spread around the curve; the two-parameter equations
support `k + s ~ 1` for a random curvature. The `covariance_structure` argument
controls whether those two random effects are allowed to correlate.

## Practical notes

- **TMB is required.** The fitting functions load a compiled C++ template; if TMB is
  not installed, the calls fail.
- **Convergence.** `fit_dd_tmb()` runs a short multi-start by default and keeps the
  best fit that passes a sanity check. A returned model carries a `converged` flag;
  when standard errors cannot be computed, the uncertainty columns come back `NA`.
- **Column names.** Whatever you call your columns via `id_var`, `x_var`, and
  `y_var`, the model frame and all method output use the canonical `id`, `x`, `y`.

## See also

- `vignette("sltb-discounting")` for the SLT-beta error model and why bounded
  indifference points need it.
- `vignette("dd-group-comparisons")` for comparing discount rates between groups.
- `vignette("delay-discounting-basics")` for single-subject fitting, `k`, and AUC.
- `vignette("bayesian-discounting")` for the same models in a Bayesian framework.
